---
title: "Pharmacoepi_Claims_Project"
format: html
editor: visual
---

# Pharmacoepidempiology: Claims Project:

## Overview:

This project will be focusing on assessing differences in treatment outcomes for Patients with Parkinson's Disease. The goal of this project is twofold:

1.  To assess if there is a significant difference in days from beginning Levodopa to a diagnosis of Parkinson's Disease with Dyskinesia between individuals using Levopdopa Monotherapy, Levodopa with Rasagiline, and Levodopa with Pramipexole.

2.  To assess if there are regional differences in prescribing frequency for our treatment arms, and if there are differences between prescriber specialty and treatment group.

As a result, I'll be splitting this code into two sections.

### Packages:

```{r}
library(tidyverse)
library(psych)
library(plotly)
library(haven)
```

### Data Import:

The data was originally sent to be as a SAS file. The form was transformed to a CSV in SAS studio prior to work. Let's import the completed CSV.

```{r}
data <- read.csv("C:\\Users\\domin\\Downloads\\dd_analytic.csv") 
```

# Aim 1:

## Summary Statistics:

Great, let's start by looking at some summary statistics:

```{r}

summary(data)
```

Ok, a quick overview:

Our total sample size is 1297 patients. The mean for rasagiline use and pramipexole use (both are binary variables) is .1164 and .118 respectively. Therefore, there are quite a few more patients on levodopa monotherapy. Approximately 23% of our total patient population is using some form of dual therapy with levodopa.

Let's confirm this number:

```{r}

# I also want to check if there are any overlapping records that are using both rasagiline or Pramipexole:

data$bothTx <- ifelse(data$rasagiline == 1 & data$pramipexole == 1, 1, 0)
table(data$bothTx)

# Ok there are 24 patients that are using both Rasagiline and Pramipexole. I want to alter the previous identification variables to remove these patients and only have them stored in the the BothTx variable:

data$rasagiline <- ifelse(data$rasagiline == 1 & data$bothTx == 0, 1, 0)
data$pramipexole <- ifelse(data$pramipexole == 1 & data$bothTx == 0, 1, 0)

# Now let's review the final numbers

table(data$rasagiline)
table(data$pramipexole)
table(data$bothTx)
```

Total Patients Dual therapy with Rasagiline: 127

Total Patients Dual therapy with Pramipexole: 129

Total Patients Therapy with all 3 groups: 24

Total Patients on any form of Dual Therapy: 304

Total Patients using Levodopa Monotherapy: 993
